Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2015-03-20 manuscript.tex
My replies to carla

Previous work by \citet{Ku1977} investigated the use of the GLQ
in gravity forward modeling.
\citet{Ku1977} numerically integrated the vertical component of the
gravitational acceleration of right rectangular prisms.
The author suggested that the accuracy of the GLQ integration depends on
the ratio between distance to the computation point and the distance between
adjacent point masses.
Based on this, \citet{Ku1977} proposed an empirical criterion that the distance
% * <berg@units.it> 2015-03-19T13:46:11.161Z:
%
%  an criterion or proposed empirical criteria
%
between adjacent point masses should be less than
the distance to the computation point.
\citet{Asgharzadeh2007} used  this criteria for the GLQ integration of
the gravity gradient tensor of tesseroids.
To our knowledge, an analysis of how well this ad hoc criteria of
\citet{Ku1977} works for gravity gradient components or for tesseroids
has never been done before.
There has also been no attempt to quantify the error committed in the GLQ
integration when applying the criteria of \citet{Ku1977}.

\citet{Li2011} devised an algorithm to automatically enforce the criteria of
\citet{Ku1977}.
Their algorithm divides the tesseroid into smaller ones instead of increasing
the number of point masses per tesseroid.
This division effectively increases the total number of point masses
because the number of points masses per tesseroid is fixed.
% * <berg@units.it> 2015-03-19T13:48:33.551Z:
%
%  sentence is repeated later.
%
A tesseroid is divided if the minimum distance to the computation point
is smaller than the largest dimension of the tesseroid.
This division is repeated recursively until all tesseroids obey the criterion.
Then, GLQ integration is performed for each of the smaller tesseroids
using the specified number of point masses.
The advantage of this adaptive discretization over
increasing the number of points masses is that the
total distribution of point masses will be greater
only close to the computation point.
This makes the adaptive discretization more computationally efficient.

\citet{Grombein2013} developed optimized formula for the gravitational fields
of tesseroids using Cartesian integral kernels.
These formulas are faster to compute and do not have singularities at the poles
like their spherical counterparts.
The Cartesian formula are numerically integrated using a Taylor series
% * <berg@units.it> 2015-03-19T13:49:40.239Z:
%
%  singular or plural?
%
expansion as per \citet{Heck2007}.
\citet{Grombein2013} use a near-zone separation to mitigate the increased error
at high latitudes.
In the so called ``near-zone'' of the computation point they use a finer
discretization composed by smaller tesseroids.
This is accomplished by dividing the tesseroids along their horizontal
dimensions.
However, the determination of an optimal size of the near-zone remains an
open question \citep{Grombein2013}.

@@

\emph{Tesseroids} consists of command-line programs
written in the C programming language.
The package includes programs to calculate
the gravitational fields of tesseroids and
rectangular prisms (in both Cartesian and spherical coordinates).
All programs receive input through
command-line arguments and the standard input channel (``STDIN'')
and output the results through the standard output channel (``STDOUT'').
For example,
the command to generate a regular grid with $NLON \times NLAT$ points,
 $g_z$ and $g_{zz}$ caused by
the tesseroids in a file ``MODELFILE'',
and save the results to a file called ``OUTPUT''
is:

@@

We will describe the stack-based algorithm
for computing the effect of a single tesseroid
on a single computation point.
The algorithm is implemented in function \emph{calc\_tess\_model\_adapt}
of the file \emph{src/lib/grav\_tess.c}.
The stack of tesseroids is represented by
the \emph{stack} variable,
an array of \emph{TESSEROID} structures.
We must define a maximum size for the stack to allocate memory for it.
Defining a maximum size allows us to
avoid an infinite loop
in case the computation point is on
(or sufficiently close to) the surface of the tesseroid.
We use the integer \emph{stktop}
to keep track of the index of
the last element in the stack (the top of the stack).

The algorithm starts by creating an empty stack of tesseroids.
Then, the stack is initialized with the input tesseroid.
% * <berg@units.it> 2015-03-19T21:04:01.768Z:
%
%  the description of the operation of stack is not clear- is it the input tesseroid array?
%
The initialization is done by copying the tesseroid into the stack
and setting \emph{stktop} to zero (the first element).
Once the stack is initialized, the steps of the algorithm are:

@@

@@

For the potential $V$, a distance-size ratio $D=1$ guarantees that the curves
for all experiments are below the $0.1\%$ error threshold.
For $g_z$, the same is achieved with $D=1.5$.
Conversely, $g_{zz}$ requires a value of $D=8$ to achieve an error level of
$0.1\%$.
For a computation height of 260 km, the error curve for $g_{zz}$
intercepts the error threshold line at $D=2.5$.
This behavior suggests that the error curves for $g_{zz}$ might depend on the
computation height.
To test this hypothesis, we computed the error curves for $g_{zz}$ at
heights 2, 10, 50, 150, and 260 km.
Figure~\ref{fig:gzz-with-height} shows the results for $g_{zz}$ at varying
computation heights.
Notice that the distance-size ratio required to achieve $0.1\%$ accuracy
decreases as the computation height increases.
% * <berg@units.it> 2015-03-19T21:07:33.445Z:
%
%  what is the height dependence due to? Should the distance not include height already? Is the height a second control parameter, next to the distance ratio? What is that due to? Is it only for gzz or also for the other gradient components? What about the gx?
%
For example, computation at 260 km height requires $D=2.5$ whereas at 10 km
height a value of $D=5.5$ is required to achieve the same accuracy.
One can take advantage of this behavior to reduce the distance-size ratio for
computations of the gravity gradient tensor at high altitudes,
saving computation time.

@@

For the gravity gradients in particular,
the distance-size ratio required for 0.1\% error decreases with height.
This can be attributed to the higher sensitivity of gravity gradients
to height variations because they decay with the distance cubed.
This behavior causes the integration kernel to have more high frequency
components, thus requiring more point masses for numerical integration.

>>>>

Previous work by \citet{Ku1977} investigated the use of the GLQ
in gravity forward modeling.
\citet{Ku1977} numerically integrated the vertical component of the
gravitational acceleration of right rectangular prisms.
The author suggested that the accuracy of the GLQ integration depends on
the ratio between distance to the computation point and the distance between
adjacent point masses.
Based on this, \citet{Ku1977} proposed an empirical criterion that the distance
% * <berg@units.it> 2015-03-19T13:46:11.161Z:
%
%  an criterion or proposed empirical criteria
%
% ^ <leouieda@gmail.com> 2015-03-20T17:41:12.218Z:
%
%  It was an criterion
%
between adjacent point masses should be less than
the distance to the computation point.
\citet{Asgharzadeh2007} used  this criteria for the GLQ integration of
the gravity gradient tensor of tesseroids.
To our knowledge, an analysis of how well this ad hoc criteria of
\citet{Ku1977} works for gravity gradient components or for tesseroids
has never been done before.
There has also been no attempt to quantify the error committed in the GLQ
integration when applying the criteria of \citet{Ku1977}.

\citet{Li2011} devised an algorithm to automatically enforce the criteria of
\citet{Ku1977}.
Their algorithm divides the tesseroid into smaller ones instead of increasing
the number of point masses per tesseroid.
% * <berg@units.it> 2015-03-19T13:48:33.551Z:
%
%  sentence is repeated later.
%
% ^ <leouieda@gmail.com> 2015-03-20T17:40:17.120Z:
%
%  I removed it.
%
A tesseroid is divided if the minimum distance to the computation point
is smaller than the largest dimension of the tesseroid.
This division is repeated recursively until all tesseroids obey the criterion.
Then, GLQ integration is performed for each of the smaller tesseroids
using the specified number of point masses.
The advantage of this adaptive discretization over
increasing the number of points masses is that the
total distribution of point masses will be greater
only close to the computation point.
This makes the adaptive discretization more computationally efficient.

\citet{Grombein2013} developed optimized formula for the gravitational fields
of tesseroids using Cartesian integral kernels.
These formulas are faster to compute and do not have singularities at the poles
like their spherical counterparts.
The Cartesian formulae are numerically integrated using a Taylor series
% * <berg@units.it> 2015-03-19T13:49:40.239Z:
%
%  singular or plural?
%
% ^ <leouieda@gmail.com> 2015-03-20T17:41:54.790Z:
%
%  It should be the plural (formulae)
%
expansion as per \citet{Heck2007}.
\citet{Grombein2013} use a near-zone separation to mitigate the increased error
at high latitudes.
In the so called ``near-zone'' of the computation point they use a finer
discretization composed by smaller tesseroids.
This is accomplished by dividing the tesseroids along their horizontal
dimensions.
However, the determination of an optimal size of the near-zone remains an
open question \citep{Grombein2013}.

@@

\emph{Tesseroids} consists of command-line programs
written in the C programming language.
The package includes programs to calculate
the gravitational fields of tesseroids and
rectangular prisms (in both Cartesian and spherical coordinates).
All programs receive input through
command-line arguments and the standard input channel (``STDIN'')
and output the results through the standard output channel (``STDOUT'').
For example,
the command to generate a regular grid with $NLON \times NLAT$ points,
 calculate $g_z$ and $g_{zz}$ caused by
the tesseroids in a file ``MODELFILE'',
and save the results to a file called ``OUTPUT''
is:

@@

The discretization algorithm is implemented in
function \emph{calc\_tess\_model\_adapt}
of the file \emph{src/lib/grav\_tess.c}.
This function calculates the effect of a single tesseroid
on a single computation point.
The stack of tesseroids is represented by
the \emph{stack} variable,
an array of \emph{TESSEROID} structures.
We must define a maximum size for the stack to allocate memory for it.
Defining a maximum size allows us to
avoid an infinite loop
in case the computation point is on
(or sufficiently close to) the surface of the tesseroid.
We use the integer \emph{stktop}
to keep track of the index of
the last element in the stack (the top of the stack).

Below, we describe the algorithm to calculate
the effect of a single tesseroid from the input model
on a single computation point.
The algorithm starts by creating an empty stack of tesseroids.
Then, the stack is initialized with the single input tesseroid.
% * <berg@units.it> 2015-03-19T21:04:01.768Z:
%
%  the description of the operation of stack is not clear- is it the input tesseroid array?
%
% ^ <leouieda@gmail.com> 2015-03-20T18:02:14.018Z:
%
%  I modified the description above and added more text below as well. Is it clearer now?
%
The initialization is done by copying the tesseroid into the stack
and setting \emph{stktop} to zero (the first element).
It is important to note that the stack is not the input tesseroid model.
Instead, it is a buffer used to temporarily store
each stage of the discretization algorithm.

Once the stack is initialized, the steps of the algorithm are:

@@

The algorithm above is repeated
for every tesseroid of the input model
and the results are summed.
This will yield the gravitational effect
of the input tesseroid model on a single point.
Thus, the computations must be repeated
for every computation point.
The whole algorithm can be summarized
in the following pseudo-code.

\begin{verbatim}
Initialize the output array with zeros.
for tesseroid in model:
  for point in grid:
    Initialize the stack with tesseroid.
    stktop = 0
    while stktop >= 0:
      Perform steps 1-6 of the algorithm.
    Sum the calculated value to the output.
\end{verbatim}

@@

For the potential $V$, a distance-size ratio $D=1$ guarantees that the curves
for all experiments are below the $0.1\%$ error threshold.
For $g_z$, the same is achieved with $D=1.5$.
Conversely, $g_{zz}$ requires a value of $D=8$ to achieve an error level of
$0.1\%$.
For a computation height of 260 km, the error curve for $g_{zz}$
intercepts the error threshold line at $D=2.5$.
This behavior suggests that the error curves for $g_{zz}$ might depend on the
computation height.
To test this hypothesis, we computed the error curves for $g_{zz}$ at
heights 2, 10, 50, 150, and 260 km.
Figure~\ref{fig:gzz-with-height} shows the results for $g_{zz}$ at varying
computation heights.
Notice that the distance-size ratio required to achieve $0.1\%$ accuracy
decreases as the computation height increases.
% * <berg@units.it> 2015-03-19T21:07:33.445Z:
%
%  what is the height dependence due to? Should the distance not include height already? Is the height a second control parameter, next to the distance ratio? What is that due to? Is it only for gzz or also for the other gradient components? What about the gx?
%
% ^ <leouieda@gmail.com> 2015-03-20T18:23:53.901Z:
%
%  The height is included in the distance and is not a control variable. This happens for all tensor components but not to the potential, gx, gy, and gz. I think this is because the decay factor for the tensor components is 1/r^3 whereas the discreatization algorithm compares D to r/size. So as the  computation point becomes closer. the field increases more rapidly than the algorithm increases the  amount of discretization. That is why you need higher D (more discretization) when you are closer to the tesseroid.
%
For example, computation at 260 km height requires $D=2.5$ whereas at 10 km
height a value of $D=5.5$ is required to achieve the same accuracy.
One can take advantage of this behavior to reduce the distance-size ratio for
computations of the gravity gradient tensor at high altitudes,
saving computation time.

@@

For the gravity gradients in particular,
the distance-size ratio required for 0.1\% error decreases with height.
We believe this is because the decay factor for
the gravity gradient components is $d^{-3}$,
whereas the discretization algorithm uses $d/L_i$.
As the computation point becomes closer to the tesseroid,
the field increases more rapidly than
the algorithm increases the amount of discretization.
Hence, a higher value of D (i.e., more discretization)
is required.

% * <leouieda@gmail.com> 2015-03-20T18:43:18.571Z:
%
%  I have modified the above paragraph to try to explain why the ratio for gzz varies with height.
%
